---
title: "Data Table basics"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## What is data.table?

data.table is an R package that provides an enhanced version of R data frames. setDT() can attach this enhanced functionality 
to a data frame without having to create a copy.

Features of the data.table package:
* fread, fwrite: fast and feature reach delimited file reader/writer functions
* low-level parallelism: many common operations (sorting, indexing, joining) are internally parallelized to use multiple CPU threads
* fast and scalable aggregations, joins, overlapping range joins, including grouping/update on join
* fast add/update/delete columns __by reference__ by group using no data copies
* fast data reshaping: dcast (pivot/wider/spread) and melt (unpivot/longer/gather)

For dtplyr (using dplyr grammar to trigger data.table operations), most recent version can be installed like this:
`devtools::install_github("tidyverse/dtplyr")`

```{r}
#devtools::install_github("tidyverse/dtplyr") # if needed
library(data.table)
```

## fread function
Fast file reader of delimited text files, with inline column selection (not all columns are loaded).
* can take an url to fetch a remote file, instead of a local file path.
* can take a linux shell command and use its output as input, e.g.:`fread("fgrep Novel junctions.csv")`
* fread can be up to 5 times faster than read.delim() when loading hundred of files (as it's the case in jaffelab::junctionCount())

```{r}
cd_dlpfc <- fread('https://dev.libd.org/vuk1tgje7i/DLPFC_bsp2_colData.gz')

head(cd_dlpfc)
```

```{r}
#class(cd_dlpfc)

selcols <- c('SAMPLE_ID', 'RNum', 'BrNum', 'Region', 'Dx', 'Age', 'Race', 'Sex', 'RIN', 'numReads', 'overallMapRate',
             'totalAssignedGene', 'mitoRate', 'rRNA_rate')
cd_dlpfc <- cd_dlpfc[, ..selcols] #special syntax for data.table, to differentiate from plain 

## can also discard unwanted column while loading the file, to save memory for large tables
cd_dg <- fread('https://dev.libd.org/vuk1tgje7i/DG_colData.gz', select = selcols, data.table=F)
class(cd_dg) # it is a data.frame
## -- transform to data.table in place (without copy)
setDT(cd_dg)
#now the class is: "data.table" "data.frame"
## it is faster than creating a copy: 
##    cd_dg <- as.data.table(cd_dg)
cd_dg
#### -- also setDF() to convert back to data.frame, but who needs that?
```
Loading row data for junctions info in these datasets
```{r}
selrcols <- c('seqnames', 'start', 'end', 'strand', 'Class', 'newGeneID', 'newGeneSymbol', 'meanExprs')
rd_dlpfc <- fread('https://dev.libd.org/vuk1tgje7i/DLPFC_bsp2_rowData.gz', select=selrcols)
# also supports col.names and colClasses parameters
rd_dg <- fread('https://dev.libd.org/vuk1tgje7i/DG_rowData.gz', select=selrcols)

## --- data.table::setnames is a nice way to rename some columns, works on data frames too
# setnames(df, oldnames, newnames)
newcolnames=c('chr', 'start', 'end', 'strand','class', 'gencodeID', 'gene', 'meanExpr')
setnames(rd_dg, selrcols, newcolnames)
setnames(rd_dlpfc, selrcols, newcolnames)

```

Other differences from data frames:
* columns of type character are not converted to factors by default
* data.table does not support row names! (during conversion from data.frame, row names can be assigned a new column) 

## Column subsetting 
```{r}

## subsetting columns:
head(rd_dg[ , c("chr", "start", "end", "strand")])
# same with quote-free syntax:
head(rd_dg[ , .(chr, strand, start, end)])
head(rd_dg[ , list(chr, strand, start, end)])
```

## Adding new columns
```{r}
## old data.frame syntax -- add a locus/location column for the junction data:

rd_dg$loc = paste0(rd_dg$chr, ':', rd_dg$start, '-', rd_dg$end,  '(',rd_dg$strand,')' )

## data.table syntax is again easier as column names are directly recognized
rd_dg[, loc:=NULL] # let's delete this column in-place
# and add it again using an easier syntax
rd_dg[, loc:= paste0(chr,':',start,'-',end,'(',strand,')') ] 

## reorder columns such that the new loc column is first
setcolorder(rd_dg, c('loc', setdiff(colnames(rd_dg), 'loc')))
head(rd_dg)

## --- for the earlier colData when we can remove and reorder multiple columns in a similar manner
## we can also remove unwanted columns in-place, without making a temporary copy for the assignment
selcols <- c('SAMPLE_ID', 'RNum', 'BrNum', 'Region', 'Dx', 'Age', 'Race', 'Sex', 'RIN', 'numReads', 'overallMapRate',
             'totalAssignedGene', 'mitoRate', 'rRNA_rate')
cd_dlpfc <- fread('https://dev.libd.org/vuk1tgje7i/DLPFC_bsp2_colData.gz')
exclude_cols=setdiff(colnames(cd_dlpfc), selcols) # difference vector: colnames(cd_dlpfc) - selcols
cd_dlpfc[, (exclude_cols) := NULL] # syntax for removing a bunch of columns at once
## same with:
##   set(cd_dlpfc, , exclude_cols, NULL) ## set function
head(cd_dlpfc)
## - set column order to how we have it in selcols
setcolorder(cd_dlpfc, selcols)


## -- back to rowdata for junctions
## - let's add loc column the same for DLPFC rowdata we loaded earlier
rd_dlpfc[, loc:= paste0(chr,':',start,'-',end,'(',strand,')') ]
## reorder columns such that the new loc column is first
setcolorder(rd_dlpfc, c('loc', setdiff(colnames(rd_dg), 'loc')))

```


##  Row filtering
```{r}
## as opposed to the data.frame syntax, column names are recognized as variables within the [ ] operator (like in subset() )
## select all novel junctions for gene SNX19
rd_dg[ class!='InGen' & gene=='SNX19' & meanExpr>0.1]


```

## General selection strategy: DT[ i, j, by ]
      DT[           i,                 j,                    by]     
           subset rows ,  select/calculate columns,     grouping  

```{r}
table(rd_dg$class)
## get all novel junctions of SNX19 with mean expression > 0.1, listed by class
rd_dg[ class!='InGen' & gene=='SNX19' & meanExpr>0.1] [order(class) ]
## order by class and decreasing meanExpr:
rd_dg[ class!='InGen' & gene=='SNX19' & meanExpr>0.1] [order(class, -meanExpr)]

## display just row counts in each group: special variable .N, 
rd_dg[ class!='InGen' & gene=='SNX19' & meanExpr>0.1, .N, by=class ]

## mean expression per novel class
rd_dg[ class!='InGen' & gene=='SNX19', .(class_mean = mean(meanExpr)), by=class]

## total expression per novel class
rd_dg[ class!='InGen' & gene=='SNX19', .(class_sum = sum(meanExpr)), by=class
      ] [order(-class_sum)]

## show multiple calculated columns for the current grouping:
## mean & total expression per novel class
rd_dg[ class!='InGen' & gene=='SNX19', 
       .(class_mean = mean(meanExpr), class_sum = sum(meanExpr)), 
          by=class]


```

Using .SD special designation : .SD = Subset of Data for the current group (a data.table with the group rows)

```{r}
rd_dg[ class!='InGen' & (gene=='SNX19'| gene=='TCF4') & meanExpr>0.1, .SD, by=.(gene, class) ]

## .SDcols can be used to limit .SD to specific columns so we can apply the same calculation function 
## to each column
## for colData metrics, calculate average values per group (Dx, Race)
cd_dlpfc[  Age<=62 & Age>=16 ,   lapply(.SD, mean), by=.(Dx, Race), 
           .SDcols=c('Age','numReads', 'totalAssignedGene', 'rRNA_rate', 'mitoRate') ][order(Dx, Race)]



```
## fast subsetting using keys
```{r}
head(rd_dg)
nrow(rd_dg)
nrow(rd_dlpfc)
#
setkey(rd_dg, loc)
setkey(rd_dlpfc, loc)
## this will create an intersection set, or "inner join": junctions shared between the two sets
nrow(merge(rd_dg, rd_dlpfc, by='loc'))

## if we want to create a "full outer join" union of two sets, placing NAs where a set misses values
nrow(merge(rd_dg, rd_dlpfc, by='loc', all=T))
head(merge(rd_dg[, .(loc, DG=meanExpr) ], 
           rd_dlpfc[, .(loc, DLPFC=meanExpr) ], by='loc', all=T))

## join
## let's pretend we don't have the loc column, so we join by: chr, start, end, strand
joincols=c('chr', 'start', 'end', 'strand')
selcols=c(joincols, 'meanExpr')
setkeyv(rd_dg, joincols, physical = F)
setkeyv(rd_dlpfc, joincols, physical = F)
mrgdt <- merge(rd_dlpfc[ , ..selcols ][, DLPFC:=meanExpr][, meanExpr:=NULL], 
           rd_dg[ , ..selcols ][,DG:=meanExpr][, meanExpr:=NULL], by=joincols, all=T)

head(mrgdt)
## this is similar to the problem of merging sets of junctions across datasets
## the junctions that had no match in the other dataset will have NA in the respective column
## so for expression levels we could set it to 0 (junction was not expressed in that dataset)
mrgdt[ is.na(mrgdt) ] = 0

mrgdt
```

## dtplyr
  R wrapper for data.table fast filtering operations to use syntax familiar to dplyr 
  Translates dplyr pipe chaining into data.table operations and chaining
  
Caveats:
  * some data.table expression have no dplyr equivalent (e.g. cross joins, rolling joins)
  * dplyr chaining involves copy operations (potential memory hazard for large tables), 
    e.g. mutate() does not modify in-place
  
```{r}
rd_dg[ class!='InGen' & (gene=='SNX19' | gene=='TCF4') , 
       .(class_mean = mean(meanExpr)), by=.(gene, class)][order(gene,class)]
## written in dtplyr:
library(dtplyr)
library(dplyr)
#rd_dg2 <- lazy_dt(rd_dg) #lazy evaluation, track operations?

rd_dg %>% filter(class!='InGen' &  (gene=='SNX19' | gene=='TCF4') ) %>%
  group_by(gene, class)  %>% summarise(class_mean = mean(meanExpr)) %>% as.data.table()

```

  
  
